Basic Model formulation

### 2015-03-03 documentation/model_formulation.Rmd
clean out and homogenize document model_formulation.Rmd

@@

The default form of the differential equation describing vegetation cover is

$$ \frac{dV}{dt} = G(V) - C(V)L $$

@@

```{r, echo=FALSE}
# Default plotting function

rho <- seq(0,1,length = 100)

@@

$$ G(V) = rV(1-\frac{V}{K}) \,,$$

with the growth rate, $r$, and the carrying capacity, $K$, decelerating the exponential growth as vegetation cover increases and reducing it to zero if $V \geq K$. Thus, it is a decelerating effect (a negative feed-back) at high vegetation cover.

@@

G <- function(P, r = 1, K = 1) r*P*(1-P/K)

@@

> The general shape of the growth function relating net growth to plant biomass, $G(V)$, is convex with a single maximum (Fig. 1). The increase in the low-biomass range expresses the increase in photosynthetic capacity with increasing leaf area. The levelling off to a plateau and the decrease in the higher range expresses self-interference effects (shading, competition); at the point $V = V_\mathrm{max}$ 'maintenance' losses equal photosynthesis and $G = 0$. Vmax is the maximum (stable) biomass in ungrazed vegetation. [...] Of the explicit forms the best known is the logistic [...] but other functions are possible.

@@

$$ r = r_0 b V^a $$

$$ G(V) = r_0 b V^{a+1} (1-\frac{V}{K}) \,,$$

@@

r <- function(P, r_0 =  1, a = 1) r_0 *P^a

@@

G_sig <- function(P, r_0 = 1, a = 1,  K = 0.9) r(P, r_0, a)*P*(1-P/K)

@@

$$ C(V) = gVL \,, $$

@@

```{r, echo=FALSE}
C <- function(V, L = 5, g = 0.01) (g*V*L)

@@

$$ C(V) = gL \,, $$

@@

```{r, echo=FALSE}
C <- function(V, L = 5, g = 0.01) g * L + V*0

@@

$$ C(V) = \frac{aVL}{1+ahV} \,,$$

@@

```{r, echo=FALSE}
C <- function(V, L = 5, a = 0.6, h = 100) (a*V*L)/(1+a*h*V)

@@

$$ C(V) = \frac{bV^{1+q}}{1+bhV^{1+q}} $$

@@

```{r, echo=FALSE}
C <- function(V, L = 5, a = 0.6, h = 100, q = 1) (a*V^(1+q)*L)/(1+a*h*(V^(1+q)) )

@@

$$ C(V) = mV + \frac{aVL}{1+ahV} \,,$$

@@

C <- function(V, m_0 = 0.05, L = 5, a = 0.6, h = 100, q = 0) (m_0 * V) +(a*V^(1+q)*L)/(1+a*h*(V^(1+q)) )

@@

The differential equation describing population cover above with $V$ and $C$ substituted would then become

$$ \frac{\mathrm{d}V}{\mathrm{d}t} = r b V (1-\frac{V}{K}) - m V - \frac{aVL}{1+ahV} $$

@@

```{r, echo=FALSE}
plot(rho,  with(parms,  m * rho + (a * rho * L)/(1 + a * h * rho)), ylab = "growth", type = "l", ylim = c(0,0.5))
lines(rho, with(parms, r* b* rho * (1 - rho/K) ), col = "green3" )

@@

odesys <- function (t, rho, parms = model_parms) {
  list( d_rho(rho, parms)  )
}

@@

@@

$$ \frac{\mathrm{d}V}{\mathrm{d}t} = r b V^{1 + \alpha} (1-\frac{V}{K}) - m V - \frac{aV^{1+q}L}{1+ahV^{1+q}} $$

@@

```{r, echo=FALSE}
plot(rho,  with(parms,  m * rho + (a * rho^( 1 + q) * L)/(1 + a * h * rho^( 1 + q))), ylab = "growth", type = "l", ylim = c(0,0.5))
  lines(rho, with(parms, r* b* rho^( 1 + alpha) * (1 - rho/K) ), col = "green3" )

@@

defparms <- list(
    m = 0.05,
    r = 1,
    b = 0.9, 
    K = 0.9, 
    a = 0.6,
    h = 100,
    L = 5, 
    alpha = 0,
    q = 0,
    c = 0,
    f = 0,
    v = 0,
    p = 0
  )
parms <- defparms
par(mar = c(4,4,2,4))
plot(rho,  with(parms,  r * b * rho * ( 1 - (rho / K)) / (1 - rho) ) , ylab = "individual growth on empty cells", type = "l", ylim = c(0,2.5), las = 1, yaxs = "i")

@@

plot(nullmodel$rho_one ~ nullmodel$time, 
     ylim = c(0,1), ylab = expression( rho["+"]), 
     xlab = "time [yr]", type = "l")

@@

And the animated lattice would look like this:
```{r, eval = FALSE, echo = FALSE, warning = FALSE}
animateCA(nullmodel, "ca_nullmodel.gif")

@@

defparms <- list(
    m = 0.05,
    r = 1,
    b = 0.9, 
    K = 0.9, 
    a = 0.6,
    h = 100,
    L = 16, 
    alpha = 0.1,
    q = 1,
    c = 0,
    f = 0,
    v = 0,
    p = 0

@@

plot(globalmodel$rho_one ~ globalmodel$time, 
     ylim = c(0,1), ylab = expression( rho["+"]), 
     xlab = "time [yr]", type = "l")

@@

## spatially-explicit models

@@

These mechanisms can be implemented in the model by using a pair-approximation approach and numerical simulations, where $G$ and $C$ can be implemented as the transition rates in a two-state spatial model depending on global and local vegetation cover.

@@

defplot(rho, G(rho, 1, set = list( f = 0.8)), ylab = "growth", col = "green3")
lines(rho,G(rho, 0.75, set = list( f = 0.8)), col = "green3")
lines(rho,G(rho, 0.5, set = list( f = 0.8)), col = "green3")
lines(rho,G(rho, 0.25, set = list( f = 0.8)), col = "green3")
lines(rho,G(rho, 0, set = list( f = 0.8)), col = "green3")

lines(rho, C(rho, set = list( L = 10)))

@@

defplot(rho, G(rho, 1, set= list(c = 0.3, b = 0.9)), ylab = "growth", col = "green3")
lines(rho,G(rho, 0.75, set= list( c = 0.3, b = 0.9)), col = "green3")
lines(rho,G(rho, 0.5, set= list( c = 0.3, b = 0.9)), col = "green3")
lines(rho,G(rho, 0.25, set= list(c = 0.3, b = 0.9)), col = "green3")
lines(rho,G(rho, 0, set= list( c = 0.3, b = 0.9)), col = "green3")

lines(rho, C(rho, set= list( L = 10)))

@@

defplot(rho, C(rho, 1, set = list( v = 0.5)), ylab = "mortality")
lines(rho,C(rho, 0.75, set = list( v = 0.5)))
lines(rho,C(rho, 0.5, set = list( v = 0.5)))
lines(rho,C(rho, 0.25, set = list( v = 0.5)))
lines(rho,C(rho, 0, set = list( v = 0.5)))

lines(rho, G(rho, set = list( b = 0.6)), col = "green3")

@@

defplot(rho, C(rho, 1, set= list(p = 0.8, L = 10)), ylab = "mortality")
lines(rho,C(rho, 0.75, set= list(p = 0.8, L = 10)))
lines(rho,C(rho, 0.5, set= list(p = 0.8, L = 10)))
lines(rho,C(rho, 0.25, set= list(p = 0.8, L = 10)))
lines(rho,C(rho, 0, set= list(p = 0.8, L = 10)))

lines(rho, G(rho, set= list(b = 0.6)), col = "green3")

@@

defplot(rho, C(rho, 1, set = list( p = 0.8, v = 0.6, L = 10, q = .5)), ylab = "mortality")
lines(rho,C(rho, 0.75, set = list( p = 0.8, v = 0.6, L = 10, q = .5)))
lines(rho,C(rho, 0.5, set = list( p = 0.8,v = 0.6,  L = 10, q = .5)))
lines(rho,C(rho, 0.25, set = list( p = 0.8,v = 0.6,  L = 10, q = .5)))
lines(rho,C(rho, 0, set = list( p = 0.8,v = 0.6,  L = 10, q = .5)))

lines(rho, G(rho, 1, set = list(c = 0.3, f = 0.9, b = 0.3, alpha = 0.5)), col = "green3")
lines(rho,G(rho, 0.75, set = list( c = 0.3,f = 0.9,  b = 0.3, alpha = 0.5)), col = "green3")
lines(rho,G(rho, 0.5, set = list(c = 0.3,f = 0.9,  b = 0.3, alpha = 0.5)), col = "green3")
lines(rho,G(rho, 0.25, set = list( c = 0.3, f = 0.9, b = 0.3, alpha = 0.5)), col = "green3")
lines(rho,G(rho, 0, set = list(c = 0.3,f = 0.9,  b = 0.3, alpha = 0.5)), col = "green3")

@@

@@

plot(spexmodel$rho_one ~ spexmodel$time, 
     ylim = c(0,1), ylab = expression( rho["+"]), 
     xlab = "time [yr]", type = "l")

@@

Thus, we can substitute $w_{1,0}$ and $w_{0,1}$ and obtain

The steady states can be simulated by solving the ODE.

parms <- defparms
parms$f = 0.9
parms$b = 0.4
parms$L = 5

attractor(parms, pairapprox = FALSE, localvals = TRUE)

@@

Visualising the attractor of the system is not as straightforward as for the models that are determined by global cover only, where the intersections of growth and mortality defined the equilibria. In the spatially-explict pair-approximation model the attractor can be visualised if we assume that clustering in the initial state is non-random, and we assign starting values for $\rho_1$ and global mortality of the population $m_1$. To do this, we need to solve the mortality function towards $\rho{1\,,1}$ to obtain initial covers of all the different pairs and singletons.

>>>>

# default plot for graphs over G or C

@@

As our starting point for model development, the default form of the differential equation describing the change of vegetation cover, $\rho$, is

$$ \frac{d\rho}{dt} = G(\rho) - C(\rho)L $$

@@

@@

$$ G(\rho) = r\rho(1-\frac{\rho}{K}) \,,$$

with the growth rate, $r$, and the carrying capacity, $K$, decelerating the exponential growth as vegetation cover increases and reducing it to zero if $\rho \geq K$. Thus, it is a decelerating effect (a negative feed-back) at high vegetation cover.

@@

G <- function(rho, r = 1, K = 1) r*rho*(1-rho/K)
rho <- seq(0,1,length = 100)

@@

> The general shape of the growth function relating net growth to plant biomass, $G(V)$, is convex with a single maximum (Fig. 1). The increase in the low-biomass range expresses the increase in photosynthetic capacity with increasing leaf area. The levelling off to a plateau and the decrease in the higher range expresses self-interference effects (shading, competition); at the point $V = V_\mathrm{max}$ 'maintenance' losses equal photosynthesis and $G = 0$. $V_\mathrm{max}$ is the maximum (stable) biomass in ungrazed vegetation. [...] Of the explicit forms the best known is the logistic [...] but other functions are possible.

@@

$$ r = r_0 b \rho^a $$

$$ G(\rho) = r_0 b \rho^{a+1} (1-\frac{\rho}{K}) \,,$$

@@

r <- function(rho, r_0 =  1, b = 1, a = 1) r_0*b *rho^a

G <- function(rho, r = 1, b =1, K = 1) r*b*rho*(1-rho/K)

@@

G_sig <- function(rho, r_0 = 1, b = 1, a = 1,  K = 0.9) r(rho, r_0, b, a)*rho*(1-rho/K)

@@

$$ C( \rho ) = g \rho L \,, $$

@@

```{r, echo=FALSE}
C <- function(rho, L = 5, g = 0.01) (g*rho*L)

@@

$$ C( \rho ) = gL \,, $$

@@

```{r, echo=FALSE}
C <- function(rho, L = 5, g = 0.01) g * L + rho*0

@@

$$ C( \rho ) = \frac{a \rho L}{1+ah \rho } \,,$$

@@

```{r, echo=FALSE}
C <- function(rho, L = 5, a = 0.6, h = 100) (a*rho*L)/(1+a*h*rho)

@@

$$ C( \rho ) = \frac{b \rho^{1+q}}{1+bh \rho^{1+q}} $$

@@

```{r, echo=FALSE}
C <- function(rho, L = 5, a = 0.6, h = 100, q = 1) (a*rho^(1+q)*L)/(1+a*h*(rho^(1+q)) )

@@

$$ C(\rho) = m \rho + \frac{a \rho L}{1+a h \rho} \,,$$

@@

C <- function(rho, m_0 = 0.05, L = 5, a = 0.6, h = 100, q = 0) (m_0 * rho) +(a*rho^(1+q)*L)/(1+a*h*(rho^(1+q)) )

@@

The differential equation describing population cover above with $G$ and $C$ substituted would then become

$$ \frac{\mathrm{d}\rho}{\mathrm{d}t} = r b \rho (1-\frac{\rho}{K}) - m \rho - \frac{a\rho L}{1+ah\rho} $$

@@

```{r, echo=FALSE}
defplot(rho,  C(rho, m = parms$m, L = parms$L, h = parms$h, a = parms$a ) )

lines(rho, G(rho, r=parms$r, K = parms$K, b=parms$b), col = "green3" )

@@

odesys <- function (t, rho, parms = parms) {
  list( d_rho(rho, parms)  )
}

@@

If we run this for low and high starting conditions, we can get the stable equilibria. Further, running an iterative series of intermediate starting conditions gives us the unstable equilibrium.

defplot(rho,  C(rho, m = parms$m, L = parms$L, h = parms$h, a = parms$a ) )

lines(rho, G(rho, r=parms$r, K = parms$K, b=parms$b), col = "green3" )

runmodel_high <- as.data.frame(ode(y = 0.99, func = odesys, times = c(1,1000), parms = parms))

points(runmodel_high[2,2],G(runmodel_high[2,2], r=parms$r, K = parms$K, b=parms$b), xpd = TRUE, pch = 20, cex = 2)

runmodel_low <- as.data.frame(ode(y = 0.0001, func = odesys, times = c(1,1000), parms = parms))

points(runmodel_low[2,2],G(runmodel_low[2,2], r=parms$r, K = parms$K, b=parms$b), xpd = TRUE, pch = 20, cex = 2)

lo <- runmodel_low[2,2]
hi <- runmodel_high[2,2]

for(i in 1:10) {

  mid <- (lo+hi)/2 
  runmodel_mid <- as.data.frame(ode(mid, func = odesys, times = c(1,1.01), parms = parms))

  if(runmodel_mid[2,2] > mid) { hi <- mid } else { lo <- mid}

}

points(mid,G(mid, r=parms$r, K = parms$K, b=parms$b), xpd = TRUE, pch = 21, cex = 1.5, bg = "white")

@@

$$ \frac{\mathrm{d}\rho}{\mathrm{d}t} = r b \rho^{1 + \alpha} (1-\frac{\rho}{K}) - m \rho - \frac{a\rho^{1+q}L}{1+ah\rho^{1+q}} $$

@@

```{r, echo=FALSE}
par(mar = c(5,4,2,3)+0.1)
plot(rho,  with(parms,  m * rho + (a * rho^( 1 + q) * L)/(1 + a * h * rho^( 1 + q))), ylab = "growth", type = "l", ylim = c(0,0.5))
  lines(rho, with(parms, r* b* rho^( 1 + alpha) * (1 - rho/K) ), col = "green3" )

@@

defparms <- list(
    m = 0.05,
    r = 1,
    b = 0.9, 
    K = 0.9, 
    a = 0.6,
    h = 100,
    L = 5, 
    alpha = 0,
    q = 0,
    c = 0,
    f = 0,
    v = 0,
    p = 0
  )

parms <- defparms
par(mar = c(4,4,2,4))
plot(rho,  with(parms,  r * b * rho * ( 1 - (rho / K)) / (1 - rho) ) , ylab = "individual growth on empty cells", type = "l", ylim = c(0,2.5), las = 1, yaxs = "i")

@@

plot(nullmodel$rho_one ~ nullmodel$time, 
     ylim = c(0,1), ylab = expression( rho["1"]), 
     xlab = "time [yr]", type = "l")

@@

And the animated lattice would look like this:

```{r, eval = FALSE, echo = FALSE, warning = FALSE}
animateCA(nullmodel, "ca_nullmodel.gif")

@@

defparms <- list(
    m = 0.05,
    r = 1,
    b = 0.9, 
    K = 0.9, 
    a = 0.6,
    h = 50,
    L = 6, 
    alpha = 0,
    q = 0,
    c = 0,
    f = 0,
    v = 0,
    p = 0

@@

plot(globalmodel$rho_one ~ globalmodel$time, 
     ylim = c(0,1), ylab = expression( rho["1"]), 
     xlab = "time [yr]", type = "l")

@@

## local feedbacks

@@

@@

parms <- defparms
parms$f = 0.8
parms$b = 0.3

attractor(parms, pairapprox = FALSE, meanfield = FALSE, localvals = TRUE)

@@

parms <- defparms
parms$c = 0.3
parms$b = 0.9

attractor(parms, pairapprox = FALSE, meanfield = FALSE, localvals = TRUE)

@@

parms <- defparms
parms$v = 0.9
attractor(parms, pairapprox = FALSE, meanfield = FALSE, localvals = TRUE)

@@

parms <- defparms
parms$p = 0.8
attractor(parms, pairapprox = FALSE, meanfield = FALSE, localvals = TRUE)

@@

parms <- defparms
parms$p = 0.8
parms$v = 0.6
parms$q = 0.5
parms$L = 6
parms$c = 0.3
parms$f = 0.9
parms$b = 0.2
attractor(parms, pairapprox = FALSE, meanfield = FALSE, localvals = TRUE)

@@

## models

These mechanisms can be modelled in different ways: The simplest is to approximate them by a mean field model, assuming that spatial structure is negligible. Second, we can use a pair-approximation model that includes the cover of pairs of cells, thereby keeping track of the local vegetation cover. Finally, numerical simulations provide the highest realism, where $G$ and $C$ can be implemented as the transition rates in a two-state spatial model depending on global and local vegetation cover.

We are developing all three approaches and compare their predictive possibilities.

Thus, we can substitute $C(\rho) = w_{1,0}$ and $G(\rho) = w_{0,1}$ and obtain

The alternative steady states and unstable equilibria can be simulated by solving the ODE.

parms <- defparms
parms$f = 0.9
parms$b = 0.4
parms$L = 5
parms$p = 0.8
parms$m = 0.01
attractor(parms, pairapprox = FALSE, localvals = TRUE)

@@

plot(spexmodel$rho_one ~ spexmodel$time, 
     ylim = c(0,1), ylab = expression( rho["1"]), 
     xlab = "time [yr]", type = "l")

@@

@@

Visualising the attractor of the system is not as straightforward as for the mean-field models that are determined by global cover only, where the intersections of growth and mortality defined the equilibria. In the spatially-explict pair-approximation model the attractor can be visualised if we assume that clustering in the initial state is non-random, and we assign starting values for $\rho_1$ and global mortality of the population $m_1$. To do this, we need to solve the mortality function towards $\rho{1\,,1}$ to obtain initial covers of all the different pairs and singletons.
